---
title: "aRchaic_view"
author: "Kushal K Dey"
date: "4/5/2017"
output: html_document
---

## aRchaic_view

`aRchaic_view ()` is an exploratory function that gives us a glimpse of the mutational patterns of a single MFF file. It produces a logo plot representation of the relative frequency of the different features - such as mutation type, flanking bases, strand, strand break and position of mutation - from the mutations recorded in the MFF file.

We demonstrate by the following example.

```{r}
library(aRchaic)
aRchaic_view(file = "../data/moderns/HG00096.mapped.ILLUMINA.bwa.GBR.low_coverage.20120522.q30.csv", output_dir = "../utilities/", title = "GBR1")
```

It says `skipping the signature aggregation step` because the `moderns.rda` file is already
present in the folder derived from `aRchaic_pool()`. If we remove the `.rda` file and rerun,
this is what we will see. 

```{r}
aRchaic_view(file = "../data/moderns/HG00096.mapped.ILLUMINA.bwa.GBR.low_coverage.20120522.q30.csv", output_dir = "../utilities/", title = "GBR1")
```
Check that here it performs signature aggregation step on just the file of interest and 
produces the logo plot in file. 

The above functions will create the following logo plot in a PNG file in the `output_dir` provided. 







Next we view an ancient MFF file from `../data/Mathieson` folder.

```{r}
aRchaic_view(file = "../data/Mathieson/I0011.1240k.subsampled.q30.csv", output_dir = "../utilities/", title = "I0011")
```

This will create the following logo plot representation.



Once can play around with the logo controls for the plot too.

```{r}
aRchaic_view(file = "../data/Mathieson/I0011.1240k.subsampled.q30.csv", output_dir = "../utilities/", title = "I0011-2", logo.control = list(ic.scale= FALSE,
                                                       title_fontsize = 20))
```

This creates the following logo plot.



Note that since `ic.scale=FALSE`, the heights of the columns in the mutation and flanking base
logo plot is the same and only the relative proportion of the different types of mutations
and flanking bases are now captured. the default however is `ic.scale=TRUE`, in which case the heights of the mutation and flanking bases are determined by the entropy.

The control parameters of the logo plot, which the user can play around with, along with their default values are the following

```{r}
logo.control.default <- list(sig_names = NULL, ic.scale=TRUE, max_pos = 20, flanking_bases=1,
         yscale_change = TRUE, xaxis=TRUE, yaxis=TRUE, xlab = " ", xaxis_fontsize=20,
         xlab_fontsize=10, title_aligner = 11, y_fontsize=20, title_fontsize = 35,
         mut_width=2, start=0.0001, renyi_alpha = 5, inflation_factor = c(2,1,2),
         pop_names=title, logoport_x = 0.25, logoport_y= 0.50, logoport_width= 0.25, logoport_height= 0.50, lineport_x = 0.9, lineport_y=0.40, lineport_width=0.32, lineport_height=0.28, breaklogoport_x = 0.94, breaklogoport_y = 0.40, breaklogoport_width=0.30, breaklogoport_height=0.45, barport_x = 0.60, barport_y=0.60, barport_width=0.25, barport_height=0.25, output_width = 1200, output_height = 700)

```

The `logoport` option fixes the position and layout of the viewport for the mutation and flanking base logo plot. The `lineport` option fixes the position and layout for the mutation trend along read plot. The `breaklogoport` option fixes the position and layout for the strand break logo plot. The `barplot` option fixes the position and layout for the strand (positive or negative strand) bar plot.

The `output_width` and `output_height` options fix the size of the logo plot image.
The inflation_factor` and `renyi_alpha` control the heights of the mutation and flanking base
logo plots if `ic.scale=TRUE`. Other options fix the axis labels, title and font sizes.

